---
title: "R Notebook"
output: html_notebook
---

---
title: "Data_Analysis_With_R"
author: "Richard Johansen"
date: "November 4, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Workshop Summary and Contact Information

**Summary:** R is a free and powerful programming language that is commonly used by researchers in both qualitative and quantitative disciplines. R provides a near comprehensive, and still expanding set of research and data analysis tools. This workshop explores the power of R for data analysis and visualization. The focus of this workshop will be hands-on exercises. No programming experience is required, but a basic comprehension of programming and statistics is benefiticial.

**Contact:**   
Email: AskData@uc.edu  
Location: 240 Braunstein Hall (GMP Library)  
Research & Data Services Website: https://libraries.uc.edu/research-teaching-support/research-data-services.html
GitHub: https://github.com/RAJohansen/UCL_Workshops
Twitter: https://twitter.com/johansen_phd

### Section I: Brief Introduction R 

##### 1. R for basic calculation
```{r}
sin(pi*15)/100
```


##### 2. R Objects & Assignment
R stores values and objects so they can be reused throughout an equation or script\
Hint alt - is a shortcut for the < - 
```{r}
x <- 1+2
y <- x +1
y
```

#### 3. Understanding functions & Getting Help in R
General recipe for functions:
```{r eval=FALSE}
function_name(argument #1 = value #1,
              argument #2 = value #2)
```

Going back to our series task, we want to create a series of numbers from 1 to 100 by 2. Luckily there are many functions already available to use in base R (many many more available from packages, which we will discuss later).\
\
Given that we are just learning R, I will tell you that the function is called "seq()"\
The first thing I do when using a new functions is to look at the documentation. You can use the ? to find R documentation.\

**HINT: Scroll to the bottom of the help page for workable examples.**\
```{r eval=FALSE}
?seq()
```

**HINT: if you can't remember exactly what function you are looking for, Use Tab.**
```{r eval=FALSE}
me<tab>
```

Additionally, if you are not sure what the function is called try a fuzzy search.\
```{r eval=FALSE}
apropos("mea") 
```


##### 1. R for basic calculation
### Section II: Exploring the Tidyverse!

#### Install and Load the tidyverse package

```{r}
require("tidyverse")
require("gapminder")
```

#### Explore the Tidyverse 
https://www.tidyverse.org/

R packages only have to be installed once but loaded everytime.\
Using require is a nice way to make sure every script has the packages needed which combines install.packages() & library()

#### 1. Basic Data Exploration
In this section we will use the gapminder data set
https://www.gapminder.org/

##### Lets assign this to gapminder
```{r}
gapminder <- gapminder
```

##### View our table
```{r eval=FALSE}
View(gapminder)
```

##### Lists the variables 
```{r}
names(gapminder)
```

##### Lets Examine the structure of the data
```{r}
str(gapminder)
```

This will become very useful when we visualize or analyze data, because we must make sure our variables are in the appropriate format!!

##### Statistical summary of the data
```{r}
summary(gapminder)
```

#### 2. Exploring our data further
**HINT: Understanding how data is indexed is crutial for R programming**

##### Lets look at column 2 
```{r}
gapminder[,2]
```

gapminder[,2]

##### Lets look at row 5
```{r}
gapminder[5,]
```


##### What value is in row 5 column 3?
```{r}
gapminder[5,3]
```


##### Based on this idea, we can make more complicated searches. Lets take the first ten observations and look at the variables:Country (1), Continent(2), Year (3), and population (5)
```{r}
gapminder[1:10,c(1:3, 5)]
```

##### What if we want to know the highest gpdPercap
```{r}
max(gapminder$gdpPercap)

```

##### Lets find the row number of the country with the highest gpdpercap
```{r}
which.max(gapminder$gdpPercap)
```

Then show me all columns for row that row

```{r}
gapminder[854,]
```


#### 2. The filter verb
The filter verb is used to look at a subset of a data set.\
Typically you combine filter with a pipe %>%

```{r}
gapminder %>% 
  filter(country == "United States")
```


##### Multiple conditions
```{r}
gapminder %>% 
  filter(year == 2007, country == "United States")
```


#### The arrange verb 
Used for sorting data by ascending or descending condition\

##### Ascending Order
```{r}
gapminder %>% 
  arrange(gdpPercap)
```


##### Descending order
```{r}
gapminder %>% 
  arrange(desc(gdpPercap))
```


##### Combining verbs
```{r}
gapminder %>% 
  filter(year == 2007) %>% 
  arrange(gdpPercap)
```


#### The mutate verb
Change or Add variables to a data set

##### Change a variable
```{r}
gapminder %>% 
  mutate(pop = pop/1000000)
```


##### Add a new variable called gdp
```{r}
gapminder %>% 
  mutate(gdp = gdpPercap * pop)
```

#### Combine all three verbs 
```{r}
gapminder %>% 
  mutate(gdp = gdpPercap * pop) %>% 
  filter(year == 2007) %>% 
  arrange(desc(gdp))
```

#### The Summarize Verb 
##### Summarize entire data set
Returns the mean of all rows (one value)
```{r}
gapminder %>% 
  summarize(meanLifeExp = mean(lifeExp))
```

##### What if we want to return the mean life exp just for 2007
```{r}
gapminder %>% 
  filter(year == 2007) %>% 
  summarize(meanLifeExp = mean(lifeExp))
```


##### Creating multiple Summaries
```{r}
gapminder %>% 
  filter(year == 2007) %>% 
  summarize(meanLifeExp = mean(lifeExp),
            totalPop = sum(pop))
```

**HINT: What data type is pop? Use str(gapminder)**

##### Convert pop to a numeric data type instead of an integer
```{r}
gapminder$pop <- as.numeric(gapminder$pop)
```


#### The group_by Verb 
The group_by verb is useful for creating aggregated groups, especially when combined with the summarize function

##### Summarize by each unique year
```{r}
gapminder %>% 
  group_by(year) %>% 
  summarize(meanLifeExp = mean(lifeExp),
            totalPop = sum(pop))
```


##### Summarize data from 2007 by continent
```{r}
gapminder %>% 
  filter(year == 2007) %>% 
  group_by(continent) %>% 
  summarize(meanLifeExp = mean(lifeExp),
            totalPop = sum(pop))
```

##### What if we want to summarize by continent over all years?
**HINT: Simply add an additional arguement to the group_by verb**
```{r}
gapminder %>% 
  group_by(year, continent) %>% 
  summarize(meanLifeExp = mean(lifeExp),
            totalPop = sum(pop))
```

#### Section III Tasks
Answer the following questions using the mtcars dataset
```{r}
mtcars <- mtcars
```

##### Task 3A. find the median mpg & wt for each group of cylinders  
```{r}
mtcars %>% 
  group_by(cyl) %>% 
  summarize(median_mpg = median(mpg),
            median_wt = median(wt))
```


##### Task 3B. find the mean of hp and drat for each group of gear and cyl & 
          #find the ratio between mean hp and mean drat
          
```{r}
mtcars %>% 
  group_by(cyl,gear) %>% 
  summarize(mean_drat = mean(drat),
            mean_hp = mean(hp),
            hp_drat_ratio = mean_hp/mean_drat)
```





















### Section III: Data Visualization
Useful resources for using base plot in R\
https://www.harding.edu/fmccown/r/\
https://www.statmethods.net/graphs/index.html\

#### 1. Default Plot
```{r}
plot(mtcars$mpg)
```

#### 2. Dotchart
```{r}
dotchart(mtcars$mpg)
```

Adding details and labels to a Simple Dotplot

```{r}
dotchart(mtcars$mpg,
         labels=row.names(mtcars),
         main="Gas Milage for Car Models", 
         xlab="Miles Per Gallon")
```

#### 3. Histogram
```{r}
hist(mtcars$mpg)
```

##### Add color and explore bin sizes

```{r}
hist(mtcars$mpg, breaks=5, col="red")
hist(mtcars$mpg, breaks=10, col="red")
hist(mtcars$mpg, breaks=15, col="red")
```

#### 4. Kernel Density Plot
First you need to save the density of the data you want to an R object\
Then plot that object using plot()
```{r}
d <- density(mtcars$mpg) # returns the density data 
plot(d) # plots the results
```

#### 5. Barplot
```{r}
barplot(mtcars$cyl)
```
HINT:: To fist create a variable called "count" to count the number of each group\
Then use the barplot() function on the object counts
```{r}
counts <- table(mtcars$cyl)
barplot(counts)
```

##### Add Chart Title and Axes

```{r}
barplot(counts, 
        main="Car Distribution", 
        xlab="Number of Gears")
```

##### Converting a Bar chart into a Stacked Bar
```{r}
counts <- table(mtcars$cyl, mtcars$gear)
barplot(counts,
        main="Car Distribution by Cylinders and Gears",
        xlab="Number of Gears",
        col = c("darkred","darkblue","orange"),
        legend = rownames(counts))

```

#### 6. Box Plots
```{r}
boxplot(mtcars$mpg~mtcars$cyl)
```

##### Boxplot with labels
```{r}
boxplot(mpg~cyl,
        data=mtcars,
        main="Car Milage Data", 
        xlab="Number of Cylinders",
        ylab="Miles Per Gallon")
```

#### 7. Pie Charts

```{r}
slices <- table(mtcars$cyl)
lbls <- c("Four", "Six", "Eight")
pie(slices,
    labels = lbls,
    main="Pie Chart of mtcars Cylindars")
```

#### 8. Scatterplot
##### Simple Scatterplot
```{r}
plot(mtcars$wt,mtcars$mpg)
```

##### Add labels
```{r}
plot(mtcars$wt, mtcars$mpg,
     main="Scatterplot Example", 
     xlab="Car Weight ",
     ylab="Miles Per Gallon ",
     pch=19)
```

#####Add linear regression line 
Regression line is (y~x) 
```{r}
plot(mtcars$wt, mtcars$mpg,
     main="Scatterplot Example", 
     xlab="Car Weight ",
     ylab="Miles Per Gallon ",
     pch=19,
     abline(lm(mtcars$mpg~mtcars$wt), col="red"))
```

#### 9. Line Graphs
```{r}
lines <- c(1:2,4,7,5,8,10,7)
plot(lines)
```

##### Graph line using blue points overlayed by a line 
```{r}
plot(lines, type="o", col="blue")
```

##### Add a title
```{r}
plot(lines, type="o", col="blue",
     main="My Line Graph")
```


#### 10. Get Inspired!!! 
##### Use the Iris data set to make a scatterplot matrix

```{r}
data("iris")
pairs(iris[1:4]) #only quantitative variables
```


##### Explore pastel theme in RColorBrewer
```{r}
require("RColorBrewer")
display.brewer.pal(3,"Pastel1") #display colorpalette
```


##### Use the function below to modify the scatterplot matrix
Put Histograms on the diagonal (from "pairs" Help)
```{r}
panel.hist  <- function(x,...)
{
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(usr[1:2], 0,1.5) )
  h <- hist(x, plot = FALSE)
  breaks <- h$breaks; nB <- length(breaks)
  y <- h$counts; y <- y/max(y)
  rect(breaks[-nB], 0, breaks[-1], y, ...)
}
```

##### Create a fancy scatterplot matrix
```{r}
pairs(iris[1:4],
      panel = panel.smooth,
      main = "Scatterplot Maxtris for Iris Data Using pairs Function",
      diag.panel = panel.hist,
      pch = 16,
      col = brewer.pal(3, "Pastel1")[unclass(iris$Species)])
```

##### Saving a plot
```{r}
jpeg('C:/temp/My_Awesome_Plot.jpg')
# Run your fancy scatter plot matrix code here!
dev.off()
```

### Section IV: Data Analysis
#### 1. Basic Stats with R
Statistics are used to summerize data!\
We use stats because it is difficult to memorize and decipher raw numbers\

#####**Example 1: Average daily car traffic for a week **
```{r}
total <- sum(5,16,15,16,13,20,25)
days <- 7
total/days
```

##### Two basic types of Statistics
**Descriptive Stats:** Uses data to describe the characteristics of a group
**Inferential Stats:** Uses the data to make predictions or draw conclusions

#### 2. Calculating descriptive statistics 
One variable
```{r}
summary(mtcars$mpg)
```

Entire Data set
```{r}
summary(mtcars)
```

Tukey's five-number summary: Min, Lower-hinge, Median, Upper-Hinge, Max (Not Labeled)
**Hint:: These five numbers are the same as a boxplot**

```{r}
fivenum(cars$mpg)
```

##### Alternative Descriptive Stats using the psych package
vars, n, mean, sd, median, trimmed, mad, min, max, range, skew, kurtosis, se
```{r}
#install.packages("psych")
library(psych)
describe(mtcars)  #vars, n, mean, sd, median, trimmed, mad, min, max, range, skew, kurtosis, se
```

##### Alternative Descriptive Stats using the pastecs pacakge
```{r}
#install.packages("pastecs")
library(pastecs)
?stat.desc()
stat.desc(mtcars)
```

#### 3. Analyzing data by groups
For this section we will use the iris dataset
```{r}
data(iris)
View(iris)
mean(iris$Petal.Width) #mean of all observation's petal.width
```

##### Split the data file and repeat analysis using "aggregate"
Allowing for the comparison of means by group
```{r}
aggregate(iris$Petal.Width ~ iris$Species, FUN = mean) # ~ means a function of...
means <- aggregate(iris$Petal.Width ~ iris$Species, FUN = mean)
plot(means)
```
**Hint:: There is significant difference between species **

##### Conducting multiple calculations at once
**Hint: The results do not keep the column headers so you need to remember the order you wrote them**
```{r}
aggregate(cbind(iris$Petal.Width, iris$Petal.Length)~ iris$Species, FUN = mean)
```

#### 4. Calculating Correlations

##### Load the mtcars dataset
```{r}
mtcars <- mtcars
```

##### Create a correlation matrix
```{r}
cor(mtcars)
```

##### Simplify the matrix to increase readability 
We can use the round() function to wrap the cor() function
```{r}
round(cor(mtcars), 2)
```

##### Correlate One pair of variables at a time
Derives r, hypothesis test, and CI\
Pearson's product-moment correlation\
```{r}
cor.test(mtcars$mpg, mtcars$wt)
```

##### Graphical Check of bivariate regression
```{r}
hist(mtcars$mpg)
hist(mtcars$wt)
plot(mtcars$wt, mtcars$mpg, abline(lm(mtcars$mpg~mtcars$wt)))
```


#### 5. Creating a Linear regression model
**Correlation:** is the strength of the association
**Regression:** is a function that can be used to predict values of another variable

##### Create a LM for miles per gallon & weight from mtcars
```{r}
reg1 <- lm(mpg~wt, data = mtcars)
reg1
```
##### Generate a summary from our LM
```{r}
summary(reg1)
```

The slope being statsitcally significant means that wt is a good predictor of mpg\
The variable weight can accounts for 0.75 or 75% of the variation in mpg\

#### 6. Calculate Multiple Regression

**Hint: Saving models as an R object allows for the extraction of additional information from model**

##### Use Six Predictors to model mpg
```{r}
reg1 <- lm(mpg ~cyl + disp + hp + wt + gear + carb, 
           data = mtcars)
```

##### Extract simple coefficents 
```{r}
reg1
```
##### Extract model details

```{r}
summary(reg1)
anova(reg1)
coef(reg1)
confint(reg1) #Confindence intervals for coefficients
resid(reg1)
hist(residuals(reg1)) #histogram of the residuals
```



