---
title: "Introduction to R"
author: "Richard Johansen"
date: "August 29, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Workshop Summary and Contact Information

**Summary:** R is a free and powerful programming language that is commonly used by researchers in both qualitative and quantitative disciplines. R provides a near comprehensive, and still expanding set of research and data analysis tools. This workshop provides a gradual introduction to the basics of programming with R using R Studio. As a participant in this workshop you will accomplish the following: explore the R Studio interface, perform basic data manipulation, learn how to install and run packages, conduct simple data analyses, and produce basic data visualizations. The focus of this workshop will be hands-on exercises to provide a deeper and more effective understanding of R. No programming experience is required, and beginners are encouraged to attend.

**Contact:**   
Email: AskData@uc.edu  
Location: 240 Braunstein Hall (GMP Library)  
Research & Data Services Website: https://libraries.uc.edu/digital-scholarship/data-services.html  
GitHub: RAJohansen  
Twitter: @DataVizJohansen  

### **R Interface Basics**

#####**R as a calculator**
```{r, eval = FALSE}
1+2
```

##### **Object and Assignments**
```{r}
x <- 1+2

y <- x +1
```

##### **Naming Conventions**
Keep names short, sweet, and relevent so anyone can follow your script!  
  
**Common Naming Conventions** 
Snake Case: "i_use_snake_case"   
Camel Case  "iUseCamelCase"  
Periods "i.use.periods"  
Combinations "iCant_decide.which_toUse"   
  
**Naming Hints**  
R is case sensative, so x != X  
Objects must start with a letter  
Use only numbers, letters, and _

##### **Understanding Functions**
Function formating: function_name(arg1 = val1, arg2 = val2)  
**HINT:** if you can't remember exactly what function you are looking we can use Tab    
Now that you found the function, How do you use it?  
**HINT:** Use the ? to find R documentation on an object  
  
###### <span style = "color:blue">**TASK 1A**</span>  
Create a sequence of numbers from 1 to 10 using the seq() function  

```{r, eval = FALSE}
?seq()

```

######<span style="color:blue">**TASK 1B**</span>  
Create a sequence of numbers from 1 to 10 by 0.5 and save it to an object called Seq  

```{r, eval = FALSE}
Seq <- seq(1,10, by =0.5) #define step by value  
Seq2 <- seq(1,10, length.out = 5) #Defined number of values  
```


























