---
title: "Mapping in R Lab"
output: html_notebook
---
<br />
**Step #1:** Load required packages <br />
HINT: These are just a few to get you started
```{r}
library(tidyverse)
library(tmap)
library(sf)
```
<br />
**Step #2:** # Import spatial object (Shapefile)
```{r}
shapefile <-  st_read("Mapping_in_R/Hands-on Lab/us_states.shp")
```
<br />
**Step #3:** Import non-spatial data frame <br />
HINT: Use the *foreign* package and the *read.dbf* function to load the census dbf file
```{r}
df <- foreign::read.dbf("Mapping_in_R/Hands-on Lab/census.dbf")
```
<br />
**Step #4:** Create identical join field on the data frame <br />
subitem Hint: To add a new column in a data frame, use the \$ and the name of the new column
```{r}
df$ST_FIPS <- df$STATE_FIPS
```
<br />
**Step #5:**Merge the spatial object and data frame using the merge function <br />
Hint: new_spatial_object <- merge(<Spatial Object>,< data frame>, by = <"join_name">)
<br />
```{r}
States_Census <- merge(shapefile,df, by = "ST_FIPS")
```
<br />
**Step #6:** Create a map of the USA using tmap <br />
HINT: For help with tmap, see examples in "Mapping in R"
```{r}
tm_shape(States_Census) +
  tm_fill(col = "MOBILEHOME", style = "cont", colorNA = NULL) + 
  tm_borders()
```

